---
title: "Identification, Inference, and Prediction"
author: "Adam M. Lauretig"
date: "12/18/2019"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Welcome
 
- Talk: Parameter identification, what it means for inference and prediction
- Latent Variable Models: Factorization Machines, Interactive Fixed Effects
- Stan code $+$ simulations!

## Who I Am

- Data Scientist
- Previously, Ph.D. in political science (at THE Ohio State University)
- Word Embeddings, Causal Inference, Bayesian Statistics
 
## Why We Model

- Descriptive Inference: is there an association between $\pmb{X}$ and $\pmb{y}$?
- Prediction: with $\pmb{X}$ and $\pmb{y}$, what is $\pmb{y}$ for a new $\pmb{X}$?
- Causal Inference: does changing $\pmb{X}$, change $\pmb{y}$?
- Shameless plug: Book Chapter **insert link**


## Parameter Identification

- Unique solution to the model
- Necessary for causal inference
- Allows for uncertainty and interpretability


## Latent Variable Models




## Slide with Plot

```{r pressure}
plot(pressure)
```

